\texttt{Supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-04-02 doc/incertae_sedis_latex2.tex
Update.

In order to apply supertree methods to rooted trees with \emph{incertae
sedis} taxa, we first propose a semantics of \emph{incertae sedis
}in terms of rooted splits.

We define rooted splits by noting that each edge of a tree divides
the tip taxa $\mathcal{L}$ into two groups: the include set $\mathcal{I}(e)$
which does not contain the root, and the exclude set $\mathcal{E}(e)$
which does contain the root. Such a split may be written
\[
\mathcal{I}(e)|\bullet\mathcal{E}(e).
\]
where the $\bullet$ indicates the root. If no taxa are \emph{incertae
sedis}, then the exclude set for a node is just the total tip set
minus the include set for the node:
\begin{align*}
\mathcal{E}(n) & =\mathcal{L}-\mathcal{I}(n).
\end{align*}

@@

Marking a taxon as \emph{incertae sedis} changes the meaning of its
sibling taxa. We seek to represent this by constructing modified splits
for each branch of the taxonomy tree. The include sets of these splits
remain unchanged, but we construct reduced exclude sets to indicate
cases where \emph{incertae sedis} taxa are allowed to intrude into
a taxon.

An \emph{incertae sedis} taxon can be moved into any of the descendants
of its siblings. Therefore, the exclude set $\mathcal{E}(n)$ for
node $n$ should not include any of its siblings marked \emph{incertae
}sedis\emph{,} but should exclude the descendants of its other siblings\emph{.}
Additionally, the exclude $\mathcal{E}(n)$ should contain the children
of all the ancestors of $n$, unless those children are \emph{incertae
sedis}.

We may also consider what happens if an \emph{incertae sedis} taxon
$A$ is placed within a sister group. Suppose $A$ is placed next
to a sibling $B$. If $B$ is also \emph{incertae sedis}, then it
should be possible to place $B$ within $A$. As a result, $A$ should
not exclude descendants of $B$ than can only be reached from $A$
by passing through a node marked \emph{incertae sedis. }

We can compute exclude sets recursively by writing the exclude set
of a node $n$ in terms of the exclude set of its parent:
\begin{align}
\mathcal{E}(n) & =\mathcal{E}(parent(n))\cup\left[\mathcal{I^{\prime}}(m)\big|m\in siblings(n)\right],\label{eq:exclude-set-formula-1}
\end{align}
where $\mathcal{I}^{\prime}(n)$ indicates the descendants of $n$
that are reachable from $n$ without passing through an \emph{incertae
sedis} node:
\begin{align*}
\mathcal{I}^{\prime}(n) & =\begin{cases}
\emptyset & \text{if }n\text{ is \emph{incertae sedis}}\\
\{n\} & \text{if }n\text{ is a leaf node}\\
\left[\mathcal{I^{\prime}}(m)\big|m\in children(n)\right] & \text{if }n\text{ is an internal node}
\end{cases}
\end{align*}
This recursion terminates if the exclude set for the root node is
set to $\emptyset$ as a boundary condition.

@@

After solving a supertree (sub) problem, we need to assign taxon names
to the supertree nodes based on the taxonomy tree in the problem.
Each taxon name $n$ corresponds to a split $S(n)=S(n)_{1}|\bullet S(n)_{2}$
on the corresponding branch of the taxonomy tree. Without \emph{incertae
sedis}, such splits are always of the form $S(n)_{1}|\bullet\mathcal{L}-S(n)_{1}$,
but with \emph{incertae sedis} taxa $S_{2}(n)$ may be smaller than
$\mathcal{L}-S(n)_{1}$.

@@

It is also possible for multiple names to apply to a single node.
For example, in Figure \ref{fig:Multiple-names-can}a, the taxon $A$
corresponds to the split $B1\,B2\,C\,|\bullet\,Y$, and is child taxon
$B$ corresponds to the split $B1\,B2\,|\bullet\,Y$. The edge leading
to $A$ is consistent with the split for $B$, but the name $B$ is
applied to the node with the smallest include group. However, in the
solution tree (Fig. \ref{fig:Multiple-names-can}b), there is is only
node for both names $A$ and $B$ to apply to. In this case, we solve
this problem by introducing a monotypic parent, and applying $A$
to the newly created parent node (Fig. \ref{fig:Multiple-names-can}c).
\begin{figure}
\hfill{}\subfloat[Taxonomy tree]{\includegraphics[scale=0.5]{Figures/names2nodes1/tax-shaded}

@@

Not all cases where multiple names apply to a single node can be resolved
in this way.

<--------------------- resume editting here ------------->

@@

\subsubsection{Multiple names fit a single node}

\paragraph{Example 1}

Suppose the taxonomy is (((B1,B2)B,C{*})A,Y) and the input tree is
(((B1,C),B2)x,Y) then the names A and B both to the node x. In this
case the names A and B are ordered.

\textbf{Case 1:} If a taxon contains 2 non-IS taxa, then it cannot
be identical with any of its children in the synthesis tree.

\textbf{Case 2:} If a taxon contains 1 non-IS taxon and $\ge1$ IS
taxa, then the taxon could be identical with is non-IS child in the
synthesis tree, if the IS taxa are placed within the child.

\textbf{Case 3:} If a node contains 0 non-IS taxa and 1 IS taxon,
then the IS taxa behaves no differently than a non-IS taxa, since
is has no siblings it could be placed into.

\textbf{Case 4:} If a node contains 0 non-IS taxa and $\ge2$ IS taxa,
then then taxon \emph{could} be identical with a non-IS child in the
synthesis tree, if all but one IS children are placed with in one
of the IS children.

\textbf{Solution:} When we assign multiple names to the same node,
then we expand the node with multiple names to have monotypic parents,
and assign the series of names to the monotypic parents. Another way
of saying this is that when a node has $\le1$ non-IS taxon and $\ge2$
taxa then the node could become monotypic by placement of the IS taxa.

\textbf{Example 2}

If ((A1,A2)A,(B1,B2)B{*},(C1,C2)C{*}); is the taxonomy with asterisks
denoting incertae sedis taxa, then the solution ((A1,A2)A,((B1,C1)mrcaB1C1,(B2,C2)mrcaB2C2)x);
has a node x that could be called B{*} or C{*}.

\textbf{Case:} If a taxa B{*} and C{*} are IS, then their tips can
be intermingled in a new clade x. Both names would then apply.

\textbf{Solution:} ??

In summary, a taxon with split $A_{1}|\bullet B_{1}$ attaches to
the most tipward node $n$ where $A_{1}\subseteq S_{1}(n)$ and $B_{1}\subseteq S_{2}(n)$.
If multiple names end up on the same node, we try to resolve the problem
by creating a monotypic node for a name that is more rootward than
all other names at that node. If this fails to resolve the problem,
we arbitrarily choose one of the names.

@@

@@

>>>>

In order to apply supertree methods to rooted trees with \emph{incertae
sedis} taxa, we first propose a semantics of \emph{incertae sedis
}in terms of rooted splits. We define rooted splits by noting that
each edge of a tree divides the tip taxa $\mathcal{L}$ into two groups:
the include set $\mathcal{I}(e)$ which does not contain the root,
and the exclude set $\mathcal{E}(e)$ which does contain the root.
Such a split may be written
\[
\mathcal{I}(e)|\bullet\mathcal{E}(e).
\]
where the $\bullet$ indicates the root. If no taxa are \emph{incertae
sedis}, then the exclude set for a node is just the total tip set
minus the include set for the node:
\begin{align*}
\mathcal{E}(n) & =\mathcal{L}-\mathcal{I}(n).
\end{align*}

@@

An \emph{incertae sedis} taxon can be moved into any of the descendants
of its siblings. Therefore marking a taxon as \emph{incertae sedis}
changes the meaning of its sibling taxa. We seek to represent this
by constructing modified splits for each branch of the taxonomy tree.
The include sets of these splits remain unchanged, but we construct
reduced exclude sets in order to allow \emph{incertae sedis} taxa
to intrude into their sibling taxa.

The reduced exclude set $\mathcal{E}(n)$ for node $n$ should therefore
not contain the descendants of its siblings marked \emph{incertae
}sedis\emph{,} but should contain the descendants of its other siblings\emph{.
}We additionally stipulate that it is not possible for \emph{incertae
sedis} taxa to intrude into their \emph{incertae sedis} siblings.
(This is not the only way to construct a split-based semantics for
\emph{incertae sedis}, but it avoids certain complications, as noted
in the discussion.) We thus define $\mathcal{I}(m,n)$ to indicate
descendants of $m$ that are excluded from $n$. Thus:
\begin{align*}
\mathcal{I}(m,n) & =\begin{cases}
\emptyset & \text{if }n\text{ is \emph{incertae sedis}\,and }m\text{ is not}\\
\mathcal{I}(m) & \text{otherwise}
\end{cases}
\end{align*}

Additionally, the exclude set $\mathcal{E}(n)$ should contain the
children of all the ancestors of $n$, unless those children are \emph{incertae
sedis}. To avoid traversing all ancestors separately for each node
$n$, we note that the excluded children of any ancestors will be
in the exclude set of the parent of $n$ already. Therefore, we can
write the exclude set of a node $n$ in terms of the exclude set of
its parent:

\begin{align}
\mathcal{E}(n) & =\mathcal{E}(parent(n))\cup\left[\mathcal{I}(m,n)\big|m\in siblings(n)\right].\label{eq:exclude-set-formula-1}
\end{align}
This formula allows us to compute exclude sets via a pre-order recursion
on the taxonomy tree. This recursion terminates if the exclude set
for the root node is set to $\emptyset$ as a boundary condition.

\textbf{Problem:} how about the goal that we should be able to get
the same outcome by moving a taxon up one level and marking it as
incertae sedis? This might require that moving A next to B, and then
moving B into A is allowed.

\textbf{Problem:} should we be able to move incertae sedis taxa down
into incertae sedis taxa? If not, then we must change the recursion.

@@

After constructing a supertree, we still need to assign taxon names
to the supertree nodes based on the taxonomy tree in the problem.
Each taxon name $n$ corresponds to a split $S(n)=S(n)_{1}|\bullet S(n)_{2}$
on the corresponding branch of the taxonomy tree. Without \emph{incertae
sedis}, such splits are always of the form $S(n)_{1}|\bullet\mathcal{L}-S(n)_{1}$,
but with \emph{incertae sedis} taxa $S_{2}(n)$ may be smaller than
$\mathcal{L}-S(n)_{1}$.

@@

It is also possible for multiple names to apply to a single node.
For example, in Figure \ref{fig:Multiple-names-can}a, the taxon $A$
corresponds to the split $B1\,B2\,C\,|\bullet\,Y$, and its child
taxon $B$ corresponds to the split $B1\,B2\,|\bullet\,Y$. The edge
leading to $A$ is consistent with the split for $B$, but the name
$B$ is applied to the node with the smallest include group. However,
in the solution tree (Fig. \ref{fig:Multiple-names-can}b), there
is is only node for both names $A$ and $B$ to apply to. In this
case, we solve this problem by introducing a monotypic parent, and
applying $A$ to the newly created parent node (Fig. \ref{fig:Multiple-names-can}c).
\begin{figure}
\hfill{}\subfloat[Taxonomy tree]{\includegraphics[scale=0.5]{Figures/names2nodes1/tax-shaded}

@@

@@

@@

If incertae sedis taxa are cannot be placed within other incertae
sedis taxa, then such an approach does not work if $y$ is already
incertae sedis. However, we take the approach that \emph{incertae
sedis} notations are not a general solution to expression of uncertain
placement.

@@

\subsubsection{Intrusion into incertae sedis taxa}

In theory it could be possible for an \emph{incertae sedis} taxon
$A$ to be placed within an \emph{incertae sedis} sibling $B$. We
initially implemented this semantics. It requires updating the recursion
to define $I(m,n)$ as the descendants of $m$ that can be access
from $n$ without traversing an \emph{incertae sedis} node.

However, if we allow this behavior by reducing the exclude sets for
$A$ and $B$, then this has the unfortunate consequence that the
members of $A$ and $B$ may be inter-digitated. In many cases, \emph{incertae
sedis} genera have siblings that are families, and the expected semantics
is that genera may be placed within their sibling families, or may
remain outside existing families. In such cases, inter-digitating
the genera would not be expected behavior. Thus, we choose the semantics
that \emph{incertae sedis} taxa exclude their \emph{incertae sedis}
siblings. This is not because the other semantics is impossible. However,
we note that use of the interdigitating semantics allows a situation
where the names for $A$ and $B$ both apply to the same node.

This semantics also complicates placement, since it is possible to
place $A$ as a sibling of $B$, and then place $B$ within $A$.
This violates the current invariant that taxa are only ever placed
within more tipward taxa. This makes the description of placement
more complicated, since we can no longer simply record all cases where
a parent node is not ancestor on the taxonomy.
